---
title: "Bayesian Optimization with TabPFN"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian Optimization with TabPFN}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

TabPFN can be used as a *surrogate model* for Bayesian optimization, following
the [PFNs4BO](https://arxiv.org/abs/2305.17535) approach (Mueller et al.,
ICML 2023). Because TabPFN predicts a full distribution over the target rather
than a single Gaussian posterior, the Expected Improvement (EI) acquisition
function can be computed in closed form from the predicted logits — and, thanks
to TabPFN's differentiable inference path, EI is differentiable with respect to
the candidate coordinates. This lets the optimizer refine its candidate points
by gradient *ascent on EI itself*.

The `rtabpfn` package exposes two functions for this:

- `tab_pfn_bayesian_optimization()` — a complete black-box optimization loop.
- `tab_pfn_bo_suggest()` — a single acquisition step for custom loops.

> Requires the full `tabpfn` package (>= 8.1.0) and
> `tabpfn-extensions[bayesian_optimization]`. Install with
> `setup_tabpfn(install_bayesian_optimization = TRUE)`.

```{r setup, include = FALSE}
library(rtabpfn)

if (requireNamespace("reticulate", quietly = TRUE)) {
  try({
    reticulate::py_run_string("import webbrowser; webbrowser.open = lambda *args, **kwargs: None")
  }, silent = TRUE)
}

has_bo <- tryCatch(check_bayesian_optimization_available(), error = function(e) FALSE)
is_ci <- Sys.getenv("CI") != "" || Sys.getenv("GITHUB_ACTIONS") == "true"
knitr::opts_chunk$set(eval = has_bo && !is_ci, message = FALSE, warning = FALSE)
```

## Minimizing a known function

Let's minimize a two-dimensional quadratic "bowl" whose minimum is known, on the
unit square defined by `bounds`:

```{r objective}
objective <- function(x) {
  (x["x1"] - 0.4)^2 + (x["x2"] - 0.6)^2   # true minimum at (0.4, 0.6), value 0
}
```

Run the Bayesian optimization loop:

```{r run-bo}
bo <- tab_pfn_bayesian_optimization(
  fun      = objective,
  bounds   = list(x1 = c(0, 1), x2 = c(0, 1)),
  n_init   = 5,
  n_iter   = 10,
  maximize = FALSE,        # we want to minimize
  device   = "auto"        # GPU if available, CPU otherwise
)

bo$best_x
bo$best_y
```

The optimizer should land near `x1 = 0.4, x2 = 0.6` with a value close to `0`.

## Tracking convergence

The returned `X` and `y` hold every point evaluated by the objective. The best
observed value so far is a running minimum:

```{r convergence, fig.width = 6, fig.height = 4}
library(ggplot2)

best_so_far <- cummin(bo$y)
history <- data.frame(iteration = seq_along(best_so_far), best = best_so_far)

ggplot(history, aes(x = iteration, y = best)) +
  geom_point() + geom_line() +
  labs(title = "Best objective value so far", x = "Evaluation", y = "f(x)") +
    theme_minimal()
```

## A custom acquisition loop

If you need fine-grained control, drive the loop yourself with
`tab_pfn_bo_suggest()`. Note that `X` must be scaled to `[0, 1]^d` and `y`
negated when minimizing:

```{r manual-loop}
set.seed(0)
d <- 2
X <- matrix(runif(4 * d), ncol = d)                 # 4 random initial points in [0,1]^2
y <- apply(X, 1, function(xi) -objective(setNames(xi, c("x1", "x2")))) # negate to maximize

for (i in 1:6) {
  x_new <- tab_pfn_bo_suggest(X, y, device = "auto")
  y_new <- -objective(setNames(x_new, c("x1", "x2")))
  X <- rbind(X, x_new)
  y <- c(y, y_new)
}

best_i <- which.max(y)
list(best_x_unscaled = X[best_i, ], best_y_negated = max(y))
```

## Notes

- The search domain is always the unit hypercube `[0, 1]^d`; the convenience
  `tab_pfn_bayesian_optimization()` scales your `bounds` to and from this space
  automatically.
- EI *maximizes*. Negate your objective when you want to minimize.
- Fits on a GPU are dramatically faster than on CPU; `device = "auto"` picks the
  GPU when one is available.
