---
title: "The Fast Endpoint and Speed Comparison"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Fast Endpoint and Speed Comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

TabPFN-3.5 ships in several variants. The **Fast** checkpoint (`3.5-fast`) is a
distilled model that delivers up to ~6x faster inference than the base `3.5`
model, while keeping the same `fit`/`predict` workflow and the same prediction
types (`numeric`, `quantiles`, `conf_int`). Select it with
`model_version = "3.5-fast"`.

This vignette benchmarks the base `3.5` and the `3.5-fast` endpoints on a small
synthetic regression problem to give a feel for the speedup.

> Requires `tabpfn >= 9.0.0`. Upgrade with `setup_tabpfn(upgrade = TRUE)`.

```{r setup, include = FALSE}
library(rtabpfn)

if (requireNamespace("reticulate", quietly = TRUE)) {
  try({
    reticulate::py_run_string("import webbrowser; webbrowser.open = lambda *args, **kwargs: None")
  }, silent = TRUE)
}

has_tabpfn <- tryCatch(check_tabpfn(), error = function(e) FALSE)
has_v35 <- tryCatch({
  numeric_version(reticulate::py_to_r(reticulate::import("tabpfn")$`__version__`)) >=
    numeric_version("9.0.0")
}, error = function(e) FALSE)

is_ci <- Sys.getenv("CI") != "" || Sys.getenv("GITHUB_ACTIONS") == "true"
knitr::opts_chunk$set(eval = has_tabpfn && has_v35 && !is_ci,
                      message = FALSE, warning = FALSE)
```

## Data

A synthetic non-linear regression problem — enough rows for the timing to be
meaningful:

```{r data}
set.seed(1)
n <- 1000
p <- 8
X <- as.data.frame(matrix(runif(n * p), ncol = p))
names(X) <- paste0("x", seq_len(p))

y <- sin(3 * X$x1) + 2 * X$x2 - 1.5 * X$x3^2 + rnorm(n, 0, 0.1)
```

## Benchmark

Time `fit()` + `predict()` for the base `3.5` model and the `3.5-fast` model.
A warm-up fit is included so the first call's one-off overhead (CUDA kernel
compilation, model load) doesn't dominate the numbers:

```{r benchmark}
benchmark_endpoint <- function(model_version) {
  # warm-up
  tab_pfn_regression(X, y, model_version = model_version, device = "auto")

  t <- system.time({
    m <- tab_pfn_regression(X, y, model_version = model_version, device = "auto")
    predict(m, X, type = "numeric")
  })
  t[["elapsed"]]
}

time_base <- benchmark_endpoint("3.5")
time_fast <- benchmark_endpoint("3.5-fast")

results <- data.frame(
  Endpoint = c("3.5 (base)", "3.5-fast"),
  Elapsed_s = round(c(time_base, time_fast), 3),
  Relative = round(c(time_base, time_fast) / time_fast, 2)
)

knitr::kable(results, caption = "fit + predict elapsed time (1000 x 8 regression)")
```

The `Relative` column expresses each endpoint's time relative to Fast
(`1.00`), so the base model's value is the observed speedup of using
`3.5-fast`.

## Predictions still match

Fast isn't just faster — it returns the same prediction types:

```{r preds}
m_fast <- tab_pfn_regression(X, y, model_version = "3.5-fast", device = "auto")

predict(m_fast, X, type = "numeric")
predict(m_fast, X, type = "quantiles", quantiles = c(0.25, 0.5, 0.75))
```

## Notes

- Timings scale with dataset size; the Fast advantage is most visible on larger
  tables and repeated fits.
- For the cloud "Thinking" endpoint (higher accuracy, point predictions only on
  the hosted API), see `tab_pfn_regression(..., thinking_mode = TRUE)`.
- A GPU is recommended for datasets past ~1000 rows.
