
```markdown
---
title: "Getting Started with rtabpfn"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with rtabpfn}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

The `rtabpfn` package provides an enhanced R interface to TabPFN, 
a foundation model for tabular data. This vignette demonstrates the 
key features, focusing on quantile predictions and uncertainty estimation.

## Setup

First, ensure the Python environment is configured:

```{r setup, eval=FALSE}
library(rtabpfn)
setup_tabpfn()
```

## Regression with Quantiles

```{r regression, eval=FALSE}
# Load data
data(mtcars)
X <- mtcars[, c("cyl", "disp", "hp", "wt")]
y <- mtcars$mpg

# Train model
model <- tab_pfn_regression(X, y)

# Get quantile predictions
preds <- predict(model, X, 
                 type = "quantiles",
                 quantiles = c(0.1, 0.5, 0.9))

head(preds)
```

## Visualizing Uncertainty

```{r visualization, eval=FALSE}
library(ggplot2)

# Combine predictions with actual values
df <- data.frame(
  actual = y,
  median = preds$.pred_q050,
  lower = preds$.pred_q010,
  upper = preds$.pred_q090
)

# Plot
ggplot(df, aes(x = actual, y = median)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  geom_errorbar(aes(ymin = lower, ymax = upper), alpha = 0.3) +
  geom_point() +
  labs(title = "Predictions with 80% Uncertainty Intervals",
       x = "Actual", y = "Predicted")
```
```
