---
title: "Time Series Forecasting"
author: "Cedric Bouffard"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Time Series Forecasting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE,
  eval = FALSE
)
library(rtabpfn)
library(dplyr)
library(lubridate)
library(ggplot2)
library(tidymodels)
```

## Introduction

The `rtabpfn` package now includes zero-shot time series forecasting capabilities through the `tabpfn-time-series` Python package. This implementation provides:

- **Zero-shot forecasting** - No training required, works out of the box
- **Probabilistic forecasts** - Native quantile predictions for uncertainty quantification
- **Point forecasts** - Both mean and median predictions
- **Automatic feature extraction** - Temporal, calendar, and seasonal features
- **Multiple time series support** - Forecast multiple series simultaneously
- **Tidymodels integration** - Compatible with tidymodels workflow

Time series forecasting with TabPFN-TS works by reframing forecasting as a tabular regression problem. The model automatically extracts relevant features and uses TabPFN's powerful foundation model to generate forecasts.

## Setup

First, ensure you have the required Python package installed:

```{r, eval=FALSE}
# Install tabpfn-time-series when setting up rtabpfn
setup_tabpfn(install_time_series = TRUE)

# Verify installation
check_time_series_available()
```

## Basic Usage

### Single Time Series

For a single time series, provide a data frame with a date column and a value column:

```{r, eval=TRUE}
# Create sample time series data
dates <- seq(as.Date("2020-01-01"), as.Date("2020-12-31"), by = "day")
trend <- seq(0, 10, length.out = length(dates))
seasonal <- sin(seq(0, 4*pi, length.out = length(dates))) * 5
noise <- rnorm(length(dates), 0, 2)
values <- trend + seasonal + noise

ts_data <- tibble(
  date = dates,
  value = values
)

head(ts_data)
```

Train the forecasting model and generate predictions:

```{r, eval=TRUE}
# Train forecasting model (using client mode by default)
model <- tab_pfn_time_series(
  train_df = ts_data,
  prediction_length = 30,
  quantiles = c(0.1, 0.5, 0.9),
  tabpfn_mode = "local",
  verbose = TRUE
)

# Generate forecasts
forecasts <- predict(model, new_data = ts_data)
head(forecasts)
```

The model automatically:
- Detects the date and value columns
- Infers the time series frequency
- Extracts temporal features (running index, calendar, seasonal)
- Generates point forecasts and quantile predictions

### Visualizing Forecasts

```{r viz-forecasts, echo=TRUE, fig.cap="Historical data with forecast"}
# Create complete example with data and forecasting
set.seed(42)
dates <- seq(as.Date("2020-01-01"), as.Date("2022-12-31"), by = "day")
trend <- seq(0, 10, length.out = length(dates))
seasonal <- sin(seq(0, 4*pi, length.out = length(dates))) * 5
noise <- rnorm(length(dates), 0, 2)
values <- trend + seasonal + noise

example_data <- tibble(
  date = dates,
  value = values
)

# Train forecasting model
model_example <- tab_pfn_time_series(
  train_df = example_data,
  prediction_length = 30,
  quantiles = c(0.1, 0.5, 0.9),
  verbose = FALSE
)

# Generate forecasts
forecasts_example <- predict(model_example, example_data)

# Prepare data for plotting
historical <- example_data %>%
  mutate(type = "historical") %>%
  rename(forecast = value)

plot_data <- forecasts_example %>%
  select(date, forecast = `0.5`) %>%
  mutate(type = "forecast")

ggplot() +
  geom_line(data = historical, aes(x = date, y = forecast, color = type), size = 0.8) +
  geom_line(data = plot_data, aes(x = date, y = forecast, color = type), 
            linetype = "dashed", size = 0.8) +
  labs(
    title = "Time Series Forecast with TabPFN",
    x = "Date",
    y = "Value",
    color = "Type"
  ) +
  scale_color_manual(values = c("historical" = "black", "forecast" = "blue")) +
  theme_minimal()
```

## Multiple Time Series

For forecasting multiple time series, include an `item_id` column:

```{r, eval=TRUE}
# Create multiple time series
set.seed(123)
dates <- seq(as.Date("2020-01-01"), as.Date("2022-12-31"), by = "day")

multi_ts_data <- tibble(
  item_id = rep(c("Product A", "Product B", "Product C"), each = length(dates)),
  date = rep(dates, 3),
  value = c(
    sin(seq(0, 4*pi, length.out = length(dates))) * 10 + rnorm(length(dates), 0, 1),
    cos(seq(0, 4*pi, length.out = length(dates))) * 10 + rnorm(length(dates), 0, 1),
    sin(seq(0, 4*pi, length.out = length(dates))) * 15 + rnorm(length(dates), 0, 2)
  )
)

# Train model
model_multi <- tab_pfn_time_series(
  train_df = multi_ts_data,
  prediction_length = 30,
  quantiles = c(0.1, 0.5, 0.9),
  item_id_col = "item_id",
  verbose = TRUE
)

# Generate forecasts
forecasts_multi <- predict(model_multi, new_data = multi_ts_data)
head(forecasts_multi)
```

## Probabilistic Forecasting

TabPFN-TS provides native support for probabilistic forecasting through quantile predictions. This is essential for risk assessment and decision-making under uncertainty.

```{r, eval=TRUE}
# Request multiple quantiles for probabilistic forecasts
model_quantiles <- tab_pfn_time_series(
  train_df = ts_data,
  prediction_length = 30,
  quantiles = c(0.05, 0.1, 0.25, 0.5, 0.75, 0.9, 0.95),
  verbose = FALSE
)

forecasts_quantiles <- predict(model_quantiles, ts_data)

# 95% prediction interval
head(forecasts_quantiles)
```

The quantile columns are named `.pred_qXXX` where XXX is the quantile as a percentage:
- `.pred_q050` - 5th percentile (lower bound)
- `.pred_q100` - 10th percentile
- `.pred_q250` - 25th percentile
- `.pred_q500` - 50th percentile (median)
- `.pred_q750` - 75th percentile
- `.pred_q900` - 90th percentile
- `.pred_q950` - 95th percentile (upper bound)

## Tidymodels Integration

The time series forecasting functionality is fully integrated with the tidymodels ecosystem, allowing you to use TabPFN-TS in workflows, tune parameters, and compare with other models.

### Basic Tidymodels Workflow

```{r, eval=TRUE}
# Create model specification
ts_spec <- tab_pfn_ts(mode = "regression") %>%
  set_engine("tabpfn_ts") %>%
  set_args(
    prediction_length = 30,
    quantiles = c(0.1, 0.5, 0.9),
    tabpfn_mode = "client",
    tabpfn_output_selection = "median"
  )

# Fit model - columns are auto-detected
ts_fit <- fit(ts_spec, data = ts_data)

# Generate predictions
tidymodels_forecasts <- predict(ts_fit, ts_data)
head(tidymodels_forecasts)
```

### Using with Workflows

```{r, eval=TRUE}
# Create a simple workflow
ts_wf <- workflow() %>%
  add_model(ts_spec)

ts_fit <- fit(ts_wf, data = ts_data)
```

### Comparing Models

```{r, eval=TRUE}
# Create different model specifications
median_spec <- tab_pfn_ts(mode = "regression") %>%
  set_engine("tabpfn_ts") %>%
  set_args(prediction_length = 30, tabpfn_output_selection = "median")

mean_spec <- tab_pfn_ts(mode = "regression") %>%
  set_engine("tabpfn_ts") %>%
  set_args(prediction_length = 30, tabpfn_output_selection = "mean")

# Fit both models (columns auto-detected)
median_fit <- fit(median_spec, data = ts_data)
mean_fit <- fit(mean_spec, data = ts_data)

# Compare forecasts
median_preds <- predict(median_fit, ts_data)
mean_preds <- predict(mean_fit, ts_data)
```

## Advanced Features

### Custom Date and Value Columns

By default, the function auto-detects date and value columns. You can explicitly specify them:

```{r, eval=TRUE}
# Custom column names
ts_data_custom <- tibble(
  datetime = as.POSIXct(dates),
  sales = values
)

model_custom <- tab_pfn_time_series(
  train_df = ts_data_custom,
  prediction_length = 30,
  date_col = "datetime",
  value_col = "sales",
  verbose = FALSE
)
```

### TabPFN Mode

Choose between client mode (default, faster, no GPU needed) or local mode (requires GPU):

```{r, eval=TRUE}
# Client mode - uses tabpfn-client API, no GPU required
model_client <- tab_pfn_time_series(
  train_df = ts_data,
  prediction_length = 30,
  tabpfn_mode = "client",
  verbose = FALSE
)

# Local mode - requires local TabPFN installation, can use GPU
# Note: Using smaller dataset for local mode to avoid memory issues
ts_data_small <- ts_data %>% slice_head(90)  # Use ~3 months for local mode
model_local <- tab_pfn_time_series(
  train_df = ts_data_small,
  prediction_length = 30,
  tabpfn_mode = "local",
  verbose = FALSE
)
```

### Output Selection

Choose between median (default) or mean for point forecasts:

```{r, eval=TRUE}
# Median predictions
model_median <- tab_pfn_time_series(
  train_df = ts_data,
  prediction_length = 30,
  tabpfn_output_selection = "median",
  verbose = FALSE
)

# Mean predictions
model_mean <- tab_pfn_time_series(
  train_df = ts_data,
  prediction_length = 30,
  tabpfn_output_selection = "mean",
  verbose = FALSE
)
```

## Function Reference

### `tab_pfn_time_series()`

Main function for creating a time series forecasting model.

```r
tab_pfn_time_series(
  train_df,
  prediction_length,
  quantiles = c(0.1, 0.5, 0.9),
  date_col = NULL,
  value_col = NULL,
  item_id_col = NULL,
  tabpfn_mode = "client",
  tabpfn_output_selection = "median",
  features = NULL,
  verbose = TRUE
)
```

**Arguments:**

- `train_df` - Data frame with training data
- `prediction_length` - Integer, number of steps to forecast
- `quantiles` - Numeric vector of quantiles for probabilistic forecasting
- `date_col` - Name of date/datetime column (NULL for auto-detection)
- `value_col` - Name of value column (NULL for auto-detection)
- `item_id_col` - Name of item_id column for multiple series
- `tabpfn_mode` - "client" (default) or "local"
- `tabpfn_output_selection` - "median" or "mean"
- `features` - List of features (NULL for automatic)
- `verbose` - Print progress information

**Returns:**

A `tab_pfn_time_series` model object that can be used with `predict()`.

### `check_time_series_available()`

Check if `tabpfn-time-series` is installed.

```r
check_time_series_available()
```

### `tab_pfn_ts()`

Tidymodels model specification for time series forecasting.

```r
tab_pfn_ts(
  mode = "regression",
  engine = "tabpfn_ts",
  prediction_length = 12,
  quantiles = c(0.1, 0.5, 0.9),
  tabpfn_mode = "client",
  tabpfn_output_selection = "median",
  ...
)
```

## Performance Considerations

- **Client mode**: Faster, no GPU required, uses TabPFN client API
- **Local mode**: Can use GPU for faster inference, requires local TabPFN installation
- **Multiple series**: Forecasting multiple series is more efficient than individual forecasts
- **Quantile computation**: More quantiles require slightly more computation time
- **Prediction length**: Longer horizons may take more time but are generally accurate

## Troubleshooting

### tabpfn-time-series Not Available

If you get an error about `tabpfn-time-series` not being available:

```r
# Check if available
check_time_series_available()

# Install if not available
setup_tabpfn(install_time_series = TRUE)

# Or install manually via reticulate
reticulate::py_install("tabpfn-time-series", pip = TRUE)
```

### Wrong Python Environment

Ensure you're using the correct Python environment:

```r
library(reticulate)

# Check current Python
py_config()

# Use your virtual environment
use_python("C:/venvs/tabpfn/Scripts/python.exe", required = TRUE)

# Verify modules are available
py_module_available("tabpfn_time_series")
```

## Example: Real-World Scenario

```{r, eval=TRUE}
# Simulate sales data with trend and seasonality (smaller dataset for local mode)
set.seed(42)
dates <- seq(as.Date("2021-01-01"), as.Date("2021-03-31"), by = "day")

# Add trend, weekly seasonality, and noise
trend <- seq(100, 150, length.out = length(dates))
weekly_seasonal <- sin(lubridate::yday(dates) / 90 * 2 * pi) * 20
noise <- rnorm(length(dates), 0, 15)

sales_data <- tibble(
  date = dates,
  sales = trend + weekly_seasonal + noise
)

# Train forecasting model (local mode - using smaller dataset)
sales_model <- tab_pfn_time_series(
  train_df = sales_data,
  prediction_length = 30,
  quantiles = c(0.1, 0.5, 0.9),
  tabpfn_mode = "local",
  verbose = TRUE
)

# Generate forecasts for the next 90 days
sales_forecasts <- predict(sales_model, sales_data)

# Analyze results
summary(sales_forecasts)
```

## References

- TabPFN-TS Paper: [arXiv:2501.02945](https://arxiv.org/abs/2501.02945)
- TabPFN-TS GitHub: https://github.com/PriorLabs/tabpfn-time-series
- Prior Labs: https://priorlabs.ai/
- rtabpfn: https://github.com/cedricbouffard/rtabpfn
